Finiteness in Cubical Type Theory

### 2020-06-25 sections/search.tex
More example on computation for search

The API for writing searches over finite domains comes from the language of the
\(\Pi\)-pretopos: with it we will show how to compose QuickCheck-like generators
for proof search, with the addition of some automation machinery that allows us
to prove things like the associativity in a couple of lines:
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{bool-assoc-auto-proof}
\end{agdalisting}

We have already, in previous sections, explored the theoretical implications of
Cubical Type Theory on our formalisation.
With this library for proof search, however, we will see two distinct
\emph{practical} applications which would simply not be possible without
computational univalence.
First and foremost: our proofs of finiteness, constructed with the API we will
describe, have all the power of full equalities.
Put another way any proof over a finite type \(A\) can be lifted to any other
type with the same cardinality.
Secondly our proof search can range over functions: we could, for instance, have
asked the prover to find if \emph{any} function over \(\AgdaDatatype{Bool}\) is
associative, and if so return it to us.
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{some-assoc}
\end{agdalisting}
The usefulness of which is dubious, but we will see a more interesting
application soon.
\section{Omniscience}
\begin{definition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience \cite{myhillErrettBishopFoundations1972} is as follows:
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Sigma {(x : A)} , P(x) \right)
  \end{equation}
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
\end{definition}
The limited principle of omniscience is non-constructive, but individual types
can themselves satisfy omniscience.
In particular, cardinal finite types are omniscient.

>>>>

The API for writing searches over finite domains comes from the language of the
\(\Pi\)-pretopos: with it we will show how to compose QuickCheck-like generators
for proof search, with the addition of some automation machinery that allows us
to prove things like the associativity in a couple of lines:
\begin{agdalisting} \label{bool-assoc-auto-proof}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{bool-assoc-auto-proof}
\end{agdalisting}

We have already, in previous sections, explored the theoretical implications of
Cubical Type Theory on our formalisation.
With this library for proof search, however, we will see two distinct
\emph{practical} applications which would simply not be possible without
computational univalence.
First and foremost: our proofs of finiteness, constructed with the API we will
describe, have all the power of full equalities.
Put another way any proof over a finite type \(A\) can be lifted to any other
type with the same cardinality.
Secondly our proof search can range over functions: we could, for instance, have
asked the prover to find if \emph{any} function over \(\AgdaDatatype{Bool}\) is
associative, and if so return it to us.
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{some-assoc}
\end{agdalisting}
The usefulness of which is dubious, but we will see a more interesting
application soon.
\section{Proof Automation And Search Techniques}
For this prover we will not resort to reflection or similar techniques: all of
the automation machinery will run in standard Agda.
To accomplish it we will leverage a common technique which we will describe
briefly here.

To understand the technique we should first notice that some proof automation
\emph{already} happens in Agda, like the following:
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{obvious}
\end{agdalisting}
While it's not a particularly impressive example, it does nonetheless
demonstrate a principle we will exploit: closed terms will compute to a normal
form.
We do not have to manually apply each computation rule or pattern match in order
for the equation to type-check.

So our task is to rewrite proof obligations like the one in
Equation~\ref{bool-assoc-auto-proof} into ones which can reduce completely. 
As it turns out, we have already described the type of proofs which can ``reduce
completely'': \emph{decidable} proofs.
If we have a decision procedure over some proposition \(P\) we can run that
decision during type checking, because the decision procedure itself is a proof
that the decision will terminate.
In code, we capture this idea with the following pair of functions:
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{is-true}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{from-true}
  \end{agdalisting}
\end{multicols}
The first is a function which derives a type from whether a decision is
successful or not.
This function is important because if we use the output of this type at any
point we will effectively force the unifier to run the decision computation.

\section{Omniscience}
First we will explore the theoretical underpinning for search, and its relation
to finiteness.
The property we are looking for is effectively the ability to exhaustively
search through some type.
Consider the proof above in Equation~\ref{bool-assoc-auto-proof}: we know the
proof could be automatic because we can check every configuration of the values
of \(x\), \(y\), and \(z\).

\begin{definition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience \cite{myhillErrettBishopFoundations1972} is as follows:
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Sigma {(x : A)} , P(x) \right)
  \end{equation}
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
\end{definition}
The limited principle of omniscience is non-constructive, but individual types
can themselves satisfy omniscience.
In particular, cardinal finite types are omniscient.
